---
title: "Initial Data Exploration"
output: html_notebook
---

# Data Exploration
Notebook to perform initial analysis on dataset contents and structure.

## Unzip data
```{r}
unzip(zipfile = "../data/raw/medical_students_dataset.zip", exdir = "../data/raw/")
```

## Load Data 
```{r}
data <- read.csv('../data/raw/medical_students_dataset.csv')
```

## Analyse Dataset Structure
There are 200,000 rows and 13 columns.

```{r}
nrow(data)
```

``` {r}
# Columns
ncol(data)
```
## Analyse First Rows
There appear to be missing values

``` {r}
head(data)
```

## Analyse Column Names/ Data Types
``` {r}
str(data)
```

## Analyse Categorical Variables

### Gender
Binary: male/female
```{r} 
unique(data$Gender)
```

### Blood.Type
Four categories: O, A, B, AB
```{r}
unique(data$Blood.Type)
```

### Diabetes
Binary: Yes/No
```{r}
unique(data$Diabetes)
```

### Smoking
Binary: Yes/No
```{r}
unique(data$Smoking)
```


# Conclusions

This dataset is very different to the accompanying description of this dataset. 
 - Rows: 200,000 > 100,000
 - Columns: Do not match descriptions.
